Sample Size Determination: Posterior Distributions Proximity

### 2024-08-21 paper/main.tex
upd paper

Through the detailed empirical study, we compare our methods with others. This comparison allows to conclude the main features of the methods we propose. Namely, that KL-sufficient sample size is typically the largest, while S-sufficient, on the contrary, is the smallest. The reason for this lies in the functions used to compare probability distributions. We encourage the reader to familiarize with the experiment code at the following link: 
\href{https://github.com/kisnikser/Posterior-Distributions-Proximity}{https://github.com/kisnikser/Posterior-Distributions-Proximity}.

@@

\subsection{Convergence Verification}

@@

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=\textwidth]{eigvals}
    \caption{Minimum eigenvalue vs available sample size}
    \label{eigvals}
\end{figure}

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=\textwidth]{synthetic-regression}
    \caption{Synthetic regression dataset}
    \label{synthetic-regression}
\end{figure}

\begin{figure}[ht]
    \centering
    \includegraphics[width=\textwidth]{liver-disorders}
    \caption{Liver Disorders dataset}
    \label{liver-disorders}
\end{figure}

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=\textwidth]{sufficient-vs-threshold}
    \caption{Sufficient sample size vs threshold}
    \label{sufficient-vs-threshold}
\end{figure}

@@

To compare our proposed methods with baselines, we used the following experiment setup. The machine learning model is linear regression. We have chosen 4 open-source datasets with regression task: Boston, Diabetes, Forestfires, and Servo. Their descriprive statistics are provided in the Table~\ref{table:descr}. We have applied 9 different baseline methods of sample size estimation on them: Lagrange Multipliers Test, Likelihood Ratio Test, Wald Test, Cross Validation, Bootstrap, Average Posterior Variance Criterion (APVC), Average Coverage Criterion (ACC), Average Length Criterion (ALC), and Utility function. Default parameters values were used for this purpose. All these methods were utilized with the help of \href{https://github.com/andriygav/SampleSizeLib}{SampleSizeLib}. As for our methods, we slightly changed the definitions of sufficiency, transferring them to terms of relative change. Namely, we consider the sample size to be sufficient if the $KL(k)$ function has a relative deviation from its value in the entire sample of no more than $\varepsilon$. Similarly with the $S(k)$ function. We fixed $\varepsilon=0.05$ and got the resulting sample sizes. This value was chosen, because other methods, especially statistical, use $0.05$ as a value of type~I error.

\begin{table}[ht]
    \centering
    \caption{Descriptive statistics of the sample sets}\label{table:descr}
    \begin{tabular}{lcc}
    \hline
        \toprule
        Sample set & Number of features, $n$ & Size of sample set, $m$ \\ 
        \midrule
        Boston Housing & 14 & 506 \\ 
        Diabetes & 20 & 576 \\ 
        Forest Fires & 13 & 517 \\ 
        Servo & 4 & 167 \\ 
        \botrule
    \end{tabular}
\end{table}

\begin{table}[ht]
    \centering
    \caption{Experimental estimation of sample size for various sample sets}\label{table:results}
    \begin{tabular}{lcccc}
    \hline
        \toprule
        Methods and sample sets & Boston & Diabetes & Forest Fires & Servo \\ 
        \midrule
        Lagrange Multipliers Test & 18 & 25 & 44 & 38 \\
        Likelihood Ratio Test & 17 & 25 & 43 & 18 \\
        Wald Test & 66 & 51 & 46 & 76 \\ 
        Cross Validation & 178 & 441 & 171 & 120 \\ 
        Bootstrap & 113 & 117 & 86 & 60 \\ 
        APVC & 98 & 167 & 351 & 20 \\ 
        ACC & 228 & 441 & 346 & 65 \\ 
        ALC & 98 & 267 & 516 & 25 \\ 
        Utility function & 148 & 172 & 206 & 105 \\ 
        \midrule
        KL (ours) & 493 & 437 & 86 & 165 \\ 
        S (ours) & 28 & 22 & 26 & 10 \\
        \botrule
    \end{tabular}
\end{table}

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=\textwidth]{dependence_on_available_sample_set.pdf}
    \caption{Estimated sample size $m^*$ versus the available sample size $m$ for each method}
    \label{dependence_on_available_sample_set}
\end{figure}

@@

Approaches to determining a sufficient sample size based on the proximity of posterior distributions of model parameters on similar subsamples are proposed. The correctness of the proposed approaches is proved under certain restrictions on the model used. The theorem on the moments of the limit posterior distribution of parameters in a linear regression model is proved. The conducted computational experiment makes it possible to analyze the properties of the proposed methods and their effectiveness.

>>>>

Through the detailed empirical study, we compare our methods with others. This comparison allows to conclude the main features of the methods we propose. Namely, that KL-sufficient sample size is typically the largest, while S-sufficient, on the contrary, is the smallest. The reason for this lies in the functions used to compare probability distributions. We encourage the reader to familiarize with the experiment code\footnote{\href{https://github.com/kisnikser/Posterior-Distributions-Proximity}{https://github.com/kisnikser/Posterior-Distributions-Proximity}}.

@@

\subsection{Convergence verification}

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=\textwidth]{eigvals}
    \caption{Minimum eigenvalue versus the available sample set size. The estimated graph, which is blue, for large sample sizes is higher than the orange graph of the power function. This asymptotic behavior of minimum eigenvalue corresponds to the theoretical results obtained.}
    \label{eigvals}
\end{figure}

\begin{figure}[ht]
    \centering
    \includegraphics[width=\textwidth]{synthetic-regression}
    \caption{Synthetic regression dataset provides results for the convergence of the proposed sample size estimation functions. The left one corresponds to the Kullback-Leibler divergence, and tends to zero, while the right one tends to one, exhibiting the behavior of s-score similarity function.}
    \label{synthetic-regression}
\end{figure}

\begin{figure}[ht]
    \centering
    \includegraphics[width=\textwidth]{liver-disorders}
    \caption{Liver Disorders dataset provides results for the convergence of the proposed sample size estimation functions. On the left, we have the Kullback-Leibler divergence, which tends towards zero as the sample size increases. On the right, we have the s-score similarity function, which tends to one as the sample size approaches infinity.}
    \label{liver-disorders}
\end{figure}

@@

@@

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=\textwidth]{sufficient-vs-threshold}
    \caption{Sufficient sample size versus the threshold parameter. S-sufficient sample size requires lower threshold values. Thus, it turns out to be more demanding of this value.}
    \label{sufficient-vs-threshold}
\end{figure}

@@

To compare our proposed methods with baselines, we used the following experiment setup. The machine learning model is linear regression. We have chosen 4 open-source datasets with regression task: Boston, Diabetes, Forestfires, and Servo. Their descriprive statistics are provided in the Table~\ref{table:descr}. We have applied 9 different baseline methods of sample size estimation on them: Lagrange Multipliers Test, Likelihood Ratio Test, Wald Test, Cross Validation, Bootstrap, Average Posterior Variance Criterion (APVC), Average Coverage Criterion (ACC), Average Length Criterion (ALC), and Utility function. They were analyzed in detail in \cite{Grabovoy2022}. Default parameters values of these methods were used. All these methods were utilized with the help of the corresponding library\footnote{\href{https://github.com/andriygav/SampleSizeLib}{https://github.com/andriygav/SampleSizeLib}}.

As for our methods, we slightly changed the definitions of sufficiency, transferring them to terms of relative change. Namely, we consider the sample size to be sufficient if the $KL(k)$ function has a relative deviation from its value in the entire sample of no more than $\varepsilon$. Similarly with the $S(k)$ function. We fixed $\varepsilon=0.05$ and got the resulting sample sizes. This value was chosen, because other methods, especially statistical, use $0.05$ as a value of type~I error.

\begin{table}[ht]
    \centering
    \caption{Descriptive statistics of the sample sets.}\label{table:descr}
    \begin{tabular}{lcc}
    \hline
        \toprule
        Sample set & Number of features, $n$ & Size of sample set, $m$ \\ 
        \midrule
        Boston Housing & 14 & 506 \\ 
        Diabetes & 10 & 576 \\ 
        Forest Fires & 13 & 517 \\ 
        Servo & 4 & 167 \\ 
        \botrule
    \end{tabular}
\end{table}

\begin{table}[ht]
    \centering
    \caption{Experimental estimation of sample size for various sample sets.}\label{table:results}
    \begin{tabular}{lcccc}
    \hline
        \toprule
        Methods and sample sets & Boston & Diabetes & Forest Fires & Servo \\ 
        \midrule
        Lagrange Multipliers Test & 18 & 25 & 44 & 38 \\
        Likelihood Ratio Test & 17 & 25 & 43 & 18 \\
        Wald Test & 66 & 51 & 46 & 76 \\ 
        Cross Validation & 178 & 441 & 171 & 120 \\ 
        Bootstrap & 113 & 117 & 86 & 60 \\ 
        APVC & 98 & 167 & 351 & 20 \\ 
        ACC & 228 & 441 & 346 & 65 \\ 
        ALC & 98 & 267 & 516 & 25 \\ 
        Utility function & 148 & 172 & 206 & 105 \\ 
        \midrule
        KL (ours) & 493 & 437 & 86 & 165 \\ 
        S (ours) & 28 & 22 & 26 & 10 \\
        \botrule
    \end{tabular}
\end{table}

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=\textwidth]{dependence_on_available_sample_set.pdf}
    \caption{Estimated sample size $m^*$ versus the available sample size $m$ for each method. The KL-divergence criterion is more conservative and requires a larger sample size. The S-sufficiency criterion, on the other hand, suggests that a minimal sample size may be sufficient.}
    \label{dependence_on_available_sample_set}
\end{figure}

@@

In this paper, we proposed methods for determining sufficient sample sizes by examining the proximity of posterior distributions of model parameters across subsamples. We proved the validity of these methods under certain model restrictions and demonstrated their effectiveness through computational experiments. While our assumption of a gaussian posterior distribution simplified the derivation of explicit formulas for the Kullback-Leibler divergence and s-score similarity function, it also limits the generalizability of our results. Future research will need to explore scenarios with non-normal posterior distributions.

Practically, we recommend calculating relative deviations from values based on the entire dataset, using small thresholds like 0.05 or 0.1. However, the computational complexity involved, especially the inversion of the covariance matrix, may limit the applicability of these methods to very large models, such as neural networks. When choosing between KL-divergence and s-score similarity, the former offers a more conservative estimate, while the latter is more optimistic. We hope this work lays the groundwork for further research, particularly in the context of more complicated machine learning models.
